$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-03-17 write_up/ms.tex
ms nearly there -- just needs to have mvt normal section completed

\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty via gradient-boosted machines}

@@

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models. Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution. Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred. Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call our convergence measure $R^*$. In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain. Since our machine learning method, gradient-boosted regression trees, provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to machine learning hyperparameter choice, and could be a complimentary additional check on MCMC convergence for applied analyses.
\end{abstract}

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic} that underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to the more recent algorithms: Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, that Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement. MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

@@

To combat this curse of hindsight, running multiple, independent, chains, which have been initialised at diverse areas of parameter space is recommended \cite{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard recommended practice \cite{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \cite{gelman2013bayesian}. If these variances are similar, $\hat{R}\approx 1$ and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \cite{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves \cite{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \cite{vehtari2019rank} and adopted \cite{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if the chains are mixed, it should not be possible to discern from a sample the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a supervised machine learning (ML) approach to measure convergence. Specifically, we train a ML model to classify the chain that caused each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across all quantiles and parameters -- any of which signal that convergence hasn't occurred. This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$ whereas $\hat{R}$ has separate values for each parameter. However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning model, we use gradient-boosted regression trees \cite{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \cite{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated and is insensitive to the GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since the outputs from GBMs are class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful status of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries and we argue provides a complementary convergence metric alongside more established measures.

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right of panel); B shows two chains that have not mixed.}
	\label{fig:marginal}
\end{figure}

@@

\begin{algorithm}[tb]
	\caption{$R^*$ calculation}
	\label{alg:R_star}
	\begin{algorithmic}
		\STATE Given $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}, S_\text{test}$:
		\FOR{$m=1$ to $N$}
		\STATE Partition, $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
		\ENDFOR
		\STATE $X_\text{train} = \{X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train}\}$
		\STATE Train ML model to classify chain id from any sample, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Predict chain id, $c^\{s\} = \text{ML}(x^\{s\}|X_\text{train})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model, $R^* = \bar{a} / (1 / N) = N/\bar{a}$
		\RETURN $R^*$
	\end{algorithmic}
\end{algorithm}

The ML model we use here is a gradient-boosted regression tree (GBM), which was introduced in \cite{friedman2001greedy}, has been shown to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s Caret package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{SK}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing. The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for models with 1000 samples) and its predictive performance was insensitive to its hyperparameters. In what follows, our hyperparameter settings were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations would be the minimum for each node and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words. For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=\bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ samples of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x^{\{s\}}|X_\text{train})\rightarrow(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^\{s\}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model, $R^{*i} = \bar{a} / (1 / N) = N/\bar{a}$
		\ENDFOR
		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
	\end{algorithmic}
\end{algorithm}

\section{Results}\label{sec:results}
To illustrate the utility of our approach, we use examples from \cite{vehtari2019rank} with a few additional cases that further illustrate the utility of $R^*$.

\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
We generate four Markov chains, where each samples from an autoregressive order 1 (AR1) process of the form,
%
\begin{equation}
X_t = \rho X_{t-1} + \epsilon_t
\end{equation}
%
where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$. Three of the chains share the same $\sigma=\bar{\sigma}$, whereas the other chain has $\sigma=1/3\bar{\sigma}$, so that it has $1/3$ of the (unconditional) variance of the others. In each replicate, we simulate 1000 iterations from all four chains and fit a gradient-boosted model to a labelled training set of 2800 randomly-chosen samples, and use it to classify samples according to the chain which generated them in an independent test set comprising the remaining 1200 observations. We perform 1000 replicates and, in each case, we report the ratio of accuracy of the predictions to the null model which assigns observations to chains uniformly at random (i.e. having an accuracy of 25\%).

In Fig. \ref{fig:ar1}A, we show how an example model fit to training data classifies observations by chain according to the sample's value. Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain. In Fig. \ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates.

%\begin{figure}[h]
%	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
%	\caption{\textbf{Autoregressive example.} A shows how the gradient-boosted model's classifications vary according to the sample's value for an example model fit; B shows the ratio of predictive accuracy to the null model accuracy across 1000 replicates; C shows the uncertainty in accuracy ratio for two series described in the text. See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
%	\label{fig:ar1}
%\end{figure}

Since the gradient-boosted models return a probability simplex for each sample, which indicates the probability it was generated by each chain, we are also able to generate uncertainty in our measure of convergence. For each sample, we randomly sample a chain classification from a multinomial distribution with class probabilities given by the outputs from the machine learning model then check whether this matches the true chain identity. The accuracy is then given by the percentage of samples where the prediction is correct. By repeating this exercise across a number of replicates, we obtain an uncertainty in the accuracy, which we convert to an accuracy ratio as before.

We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the "unconverged data"); and another, where all chains sample from the same distribution (we call this the "converged data"). In Fig. \ref{fig:ar1}C, we show the uncertainty distributions for the accuracy ratio that result from sampling from the underlying fit of a single gradient-boosted model to each dataset. For the unconverged data, the distribution has its bulk of mass away from 1, indicating lack of convergence. For the converged data, the distribution is centred on 1, indicating convergence.

\subsection{Lack of convergence of the joint distribution: multivariate normal model}\label{sec:multivariate_normal}
In this example, we use the 250-dimensional multivariate normal target, where its precision matrix is generated from a Wishart distribution \cite{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for X iterations: at this point, all the $\hat{R}<1.01$ and the effective sample size for each dimension exceeded 400 (recommendations from \cite{vehtari2019rank}). Using these samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig. X.

\subsection{Infinite variance: Cauchy example}
We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance. Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
%
\begin{equation}
x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50,
\end{equation}
%
and monitor convergence of each $x_j$ separately. We call this parameterisation the ``nominal'' version of this model.

@@

In Fig. \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations. As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, the samples still contain information about the chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$. Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig. \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5). In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has been reached. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.

%\begin{figure}[h]
%	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
%	\caption{\textbf{Cauchy example.} A shows 1000 samples from the $R^*$ distribution generated under two model parameterisations. See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
%	\label{fig:cauchy}
%\end{figure}

\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
In this example, we examine a simple classic example used to highlight difficulties in performing sampling for hierarchical models: referred to as the "Eight schools" model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.

>>>>

\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines}

@@

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models. Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution. Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred. Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call our convergence measure $R^*$. In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distributions, which is currently largely uncovered by existing approaches. Since our machine learning method, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
\end{abstract}

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement. MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

@@

To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \cite{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \cite{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \cite{gelman2013bayesian}. If these variances are similar, $\hat{R}\approx 1$, and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \cite{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\hat{R}$ \cite{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \cite{vehtari2019rank} and adopted \cite{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a supervised machine learning (ML) approach to measure convergence. Specifically, we train a ML model to classify the chain that generated each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent approaches. This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$, whereas $\hat{R}$ has separate values for each parameter. However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning model, we use gradient-boosted regression trees \cite{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \cite{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to train then test). It is also insensitive to the GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since the outputs from GBMs are class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (to the right of panel); B shows two chains that have not mixed.}
	\label{fig:marginal}
\end{figure}

@@

\begin{algorithm}[tb]
	\caption{$R^*$ calculation}
	\label{alg:R_star}
	\begin{algorithmic}
		\STATE Given chain-wise samples from the target, $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}$ and a test set length, $S_\text{test}$:
		\FOR{$m=1$ to $N$}
		\STATE Create train and test sets by random-sampling (w/o replacement), $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
		\ENDFOR
		\STATE Stack $X_\text{train} = (X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train})^T$
		\STATE Stack $X_\text{test} = (X^{\{1\}}_\text{test},X^{\{2\}}_\text{test},...,X^{\{N\}}_\text{test})^T$
		\STATE Train ML model to classify chain id from any sample, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id, $c^{\{s\}} = \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^* = \bar{a} / (1 / N) = N/\bar{a}$
		\RETURN $R^*$
	\end{algorithmic}
\end{algorithm}

The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing. The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for models with 1000 samples), and its predictive performance was insensitive to its hyperparameters. In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words. For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N/\bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ samples of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} / (1 / N) = N/\bar{a}$
		\ENDFOR
		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
	\end{algorithmic}
\end{algorithm}

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} with one additional example: the multivariate normal model described in \cite{hoffman2014no}.

\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
We generate four Markov chains, where each samples from an autoregressive order 1 (AR1) process of the form,
%
\begin{equation}
X_t = \rho X_{t-1} + \epsilon_t
\end{equation}
%
where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$ and $t=1,2,...,1000$. Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others. In Fig. \ref{fig:ar1}A, we show how a GBM model fitted to these data classifies observations according to a sample's value. Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.

To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e. with one chain with a lower variance). We then fit a gradient-boosted model to a labelled training set of 2800 randomly chosen samples. The fitted model is then used to classify samples in an independent test set (comprising the remaining 1200 observations) according to the chain which generated them. For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}. In Fig. \ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates.

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the sample's value for an example model fit; B shows the calculated $R^*$ values across 1000 replicates; C shows the $R^*$ distributions for the two series described in the text. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:ar1}
\end{figure}

Since GBMs return a probability simplex for each sample, which, for each chain, indicates the probability that a sample was generated by it, we can also generate a measure of uncertainty in $R^*$. Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done. We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data'); and another, where all chains sample from the same distribution (we call this the ``converged'' data). In Fig. \ref{fig:ar1}C, we show the $R^*$ distributions in each case. For the unconverged data, the distribution has its bulk of mass away from 1, indicating lack of convergence. For the converged data, the distribution is centred on 1, indicating convergence.

\subsection{Lack of convergence of the joint distribution: multivariate normal model}\label{sec:multivariate_normal}
In this example, we use the 250-dimensional multivariate normal target, where its precision matrix is generated from a Wishart distribution \cite{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for three different iteration counts: 400, 2000 and 5000. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 5000 iteration case, the $R^*$ distribution remains stubbornly shifted somewhat rightwards of $R^*=1$ (coincidently, the $\hat{R}$ values ...).

\begin{figure}[h]
	\centerline{\includegraphics[width=0.6\textwidth]{../output/mvt_three.pdf}}
	\caption{\textbf{Multivariate normal example.} Plot shows $R^*$ distributions obtained for three different MCMC sample sizes. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:mvt}
\end{figure}

\subsection{Infinite variance: Cauchy example}
We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance. Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
%
\begin{equation}
x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50,
\end{equation}
%
We call this parameterisation the ``nominal'' version of this model.

@@

In Fig. \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations. As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$. Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig. \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5). In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has been reached. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
	\caption{\textbf{Cauchy example.} A shows 1000 MCMC samples from the $R^*$ distribution generated under two model parameterisations; B shows the same but with 10,000 MCMC samples used (which are then thinned by a factor of 5). See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:cauchy}
\end{figure}

\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight school'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.

The model can be parameterised two ways, as described in \cite{vehtari2019rank}. The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
%
\begin{align*}
\theta_j &\sim \N(\mu, \tau) \\
y_j &\sim \N(\theta_j, \sigma_j).
\end{align*}
%
The ``non-centered'' parameterisation recodes this model in a way which does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$. This can be written as,
%
\begin{align*}
\tilde{\theta}_j &\sim \N(0, 1) \\
\theta_j &= \mu + \tau \tilde{\theta}_j \\
y_j &\sim \N(\theta_j, \sigma_j).
\end{align*}
%
In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
of the distribution of these effects. In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.
